### 2022-03-29 pre/pre.tex
db

@@

The standard resolution and superposition calculi do not work directly on formulas, but
on their normal forms, {\em clauses}. Thus, the initial problem, expressed as a
set of formulas, must be transformed into a set of clauses. For propositional
and first-order logic, there exists such a transformation that does not affect
satisfiability of the problem \cite{nw-01-small-cnf}. In higher-order logic,
formulas are first-class citizens and higher-order calculi can in general work
with nonnormalized problem, in its original form. However, resolution and
superposition-based higher-order calculi still perform best when the problem is clausified
using transformation similar to the one used in the first-order case. We have analyzed
different approaches to clausification in higher-order context \cite{bbtv-21-full-ho-sup}.

>>>>

In practice, many implementations of $\lambda$-terms avoid named representation
of bound variables described above by using \emph{locally nameless
representation} \cite{ac-12-locally-nameless}. In this representation, free
variables retain their names, while abstractions are simply represented by a
$\lambda$ symbol, without any names. Further, bound variables are represented
using \emph{De Bruijn indices}. A bound variable with De Bruijn index
$\dbvar{i}$ is bound by the $i+1$-th enclosing $\lambda$-binder. For example,
the term $\lambda x. \cst{f} \, (\lambda y z. \cst{g} \, y \, x)$ is represented
as $\lambda\, \cst{f} \, (\lambda \lambda\, \cst{g} \, \db{0} \, \db{2})$.
Locally nameless representation reduces the $\alpha$-equivalence test to 
the syntactic equality test.

@@

The standard resolution and superposition calculi do not work directly on formulas, but
on normal forms, {\em clauses}. Thus, the initial problem, expressed as a
set of formulas, must be transformed into a set of clauses. For propositional
and first-order logic, there exists such a transformation that does not affect
satisfiability of the problem \cite{nw-01-small-cnf}. In higher-order logic,
formulas are first-class citizens and higher-order calculi can in general work
with nonnormalized problem, in its original form. However, resolution and
superposition-based higher-order calculi still perform best when the problem is clausified
using a transformation similar to the one used in the first-order case. We have analyzed
different approaches to clausification in a higher-order context \cite{bbtv-21-full-ho-sup}.

### 2022-03-29 unif/unif.tex
more
db

\emph{Parameters} and \emph{body} for any term $\param[\seq x]s$ are defined to
be $\seq x$ and $s$ respectively, where $s$ is not a $\lambda$-abstraction. The
\emph{size} of a term is inductively defined as $\text{size}(F) = 1$;
$\text{size}(x) = 1$; $\text{size}(\cst{f}) = 1$; $\text{size}(s\,t) =
\text{size}(s) + \text{size}(t)$; $\text{size}(\lambda x.\, s) = \text{size}(s)
+ 1$. A term is in \emph{head normal form} ({\em hnf}) if it is of the form
$\lambda \overline{x}.\,a\,\overline{t}$, where $a$ is a free variable, a bound
variable, or a constant. In this case, $a$ is called the \emph{head} of the
term. Note that this relaxes the condition that the term needs to be in
$\beta$-normal form to determine its head. A term is called flex or rigid if its
head is flex or rigid, respectively. By $\nf{s}{\textsf{h}}$ we denote the term
obtained from a term $s$ by repeated $\beta$-reduction of the leftmost outermost
redex until it is in hnf. Unless stated otherwise, we view terms syntactically,
as opposed to $\alpha\beta\eta$-equivalence classes. The common context
$\mathcal{C}(s,t)$ of two $\eta$-long $\beta$-reduced terms $s$~and~$t$ of the
same type is defined inductively as follows, assuming that $a \not= b$:
$\mathcal{C}(\lambda x.\, s, \lambda y.\, t) = \lambda x.\,
\mathcal{C}(s,\{y\mapsto x\}t)$; $\mathcal{C}(a\,\overline{s}_m,
b\,\overline{t}_n) = \square$; $\mathcal{C}(a\,\overline{s}_m,
a\,\overline{t}_m) = a\,\mathcal{C}(s_1,t_1)\,\ldots\,\mathcal{C}(s_m,t_m)$.
Unless otherwise stated, we take unification constraint $s \unif t$ to be an
unordered pair of two terms of the same type. To ease notation, we do not write
parentheses around application of substitutions to terms (or other objects containing terms);
in other words we shorten $\sigma(\theta(\varrho(s)))$ to $\sigma\theta\varrho \, s$.

@@

The fundamental idea of first-order fingerprint indexing carries over to
higher-order terms---application of a substitution does not change the rigid skeleton of a term.
However, to extend fingerprint indexing to
higher-order terms, we must address the issues of $\alpha\beta\eta$-normalization and figure out how to cope with $\lambda$-abstractions and
bound variables. To that end, we define a function $\hotofo{t}$, defined on
%TODO: @PETAR \eta-long?
% Yes
$\eta$-long $\beta$-reduced terms in De Bruijn \cite{db-75-dbindices} notation:
%TODO: @PETAR: also mention that this is an encoding into FOL before the definition?
% @ALEX: There is a sentence about that later. I do not see how it can easily fit in this description :(
\begin{align*}
& \hotofo{F \, \overline{s}} = F
&& \hotofo{\dbvar{i} \, \overline{s}_n } = \cstdba{i}(\hotofo{s_1}, \ldots, \hotofo{s_n})
&& \hotofo{\cst{f} \, \overline{s}_n } = \cst{f}(\hotofo{s_1}, \ldots, \hotofo{s_n}) &&
\hotofo{\lambda \overline{x}. \, s  } = \hotofo{s}
\end{align*}
\looseness=-1
We let $\dbvar{i}$ be a bound variable of type $\alpha$ with De Bruijn index $i$,
and $\cstdba{i}$ be a fresh constant corresponding to this variable. All constants
$\cstdba{i}$ must be fresh. 
Effectively, $\hotofo{\;}$ transforms an $\eta$-long $\beta$-reduced higher-order term to an untyped first-order term. Let $\nf{t}{\beta\eta}$ be
the $\eta$-long $\beta$-reduced form of $t$; the higher-order generic
fingerprinting function $\text{gfpf}_\textsf{ho}$, which relies on conversion $\nametodb{t}$ from
named to De Bruijn representation, is defined as
\[
  \text{gfpf}_\textsf{ho}(t,p) =
    \text{gfpf}(\hotofo{\nametodb{\nf{t}{\beta\eta}}}, p)
\]

>>>>

\emph{Parameters} and \emph{body} for any term $\param[\seq x]s$ are defined to
be $\seq x$ and $s$ respectively, where $s$ is not a $\lambda$-abstraction. The
\emph{size} of a term is inductively defined as $\text{size}(F) = 1$;
$\text{size}(x) = 1$; $\text{size}(\cst{f}) = 1$; $\text{size}(s\,t) =
\text{size}(s) + \text{size}(t)$; $\text{size}(\lambda x.\, s) = \text{size}(s)
+ 1$. A term is in \emph{head normal form} ({\em hnf}) if it is of the form
$\lambda \overline{x}.\,a\,\overline{t}$, where $a$ is a free variable, a bound
variable, or a constant. In this case, $a$ is called the \emph{head} of the
term. Note that this relaxes the condition that the term needs to be in
$\beta$-normal form to determine its head. A term is called flex or rigid if its
head is flex or rigid, respectively. By $\nf{s}{\textsf{h}}$ we denote the term
obtained from a term $s$ by repeated $\beta$-reduction of the leftmost outermost
redex until it is in hnf. Unless stated otherwise, we view terms syntactically,
as opposed to $\alpha\beta\eta$-equivalence classes. The common context
$\mathcal{C}(s,t)$ of two $\eta$-long $\beta$-reduced terms $s$~and~$t$ of the
same type is defined inductively as follows, assuming that $a \not= b$:
$\mathcal{C}(\lambda x.\, s, \lambda y.\, t) = \lambda x.\,
\mathcal{C}(s,\{y\mapsto x\}t)$; $\mathcal{C}(a\,\overline{s}_m,
b\,\overline{t}_n) = \square$; $\mathcal{C}(a\,\overline{s}_m,
a\,\overline{t}_m) = a\,\mathcal{C}(s_1,t_1)\,\ldots\,\mathcal{C}(s_m,t_m)$.
Unless otherwise stated, we take unification constraint $s \unif t$ to be an
unordered pair of two terms of the same type. To ease notation, we do not write
parentheses around application of substitutions to terms (or other objects containing terms);
in other words we shorten $\sigma(\theta(\varrho(s)))$ to $\sigma\theta\varrho \, s$. When we write $\tau_1
\rightarrow \cdots \rightarrow \tau_n \rightarrow \upsilon$ we assume $\upsilon$
to be a base type, by convention.

@@

The fundamental idea of first-order fingerprint indexing carries over to
higher-order terms---application of a substitution does not change the rigid skeleton of a term.
However, to extend fingerprint indexing to
higher-order terms, we must address the issues of $\alpha\beta\eta$-normalization and figure out how to cope with $\lambda$-abstractions and
bound variables. To that end, we define a function $\hotofo{t}$, defined on
%TODO: @PETAR \eta-long?
% Yes
$\eta$-long $\beta$-reduced terms in De Bruijn \cite{db-75-dbindices} notation:
%TODO: @PETAR: also mention that this is an encoding into FOL before the definition?
% @ALEX: There is a sentence about that later. I do not see how it can easily fit in this description :(
\begin{align*}
& \hotofo{F \, \overline{s}} = F
&& \hotofo{\db{i} \, \overline{s}_n } = \cstdba{i}(\hotofo{s_1}, \ldots, \hotofo{s_n})
&& \hotofo{\cst{f} \, \overline{s}_n } = \cst{f}(\hotofo{s_1}, \ldots, \hotofo{s_n}) &&
\hotofo{\lambda \overline{x}. \, s  } = \hotofo{s}
\end{align*}
\looseness=-1
We let $\db{i}$ be a bound variable of type $\alpha$ with De Bruijn index $i$,
and $\cstdba{i}$ be a fresh constant corresponding to this variable. All constants
$\cstdba{i}$ must be fresh. 
Effectively, $\hotofo{\;}$ transforms an $\eta$-long $\beta$-reduced higher-order term to an untyped first-order term. Let $\nf{t}{\beta\eta}$ be
the $\eta$-long $\beta$-reduced form of $t$; the higher-order generic
fingerprinting function $\text{gfpf}_\textsf{ho}$, which relies on conversion $\nametodb{t}$ from
named to De Bruijn representation, is defined as
\[
  \text{gfpf}_\textsf{ho}(t,p) =
    \text{gfpf}(\hotofo{\nametodb{\nf{t}{\beta\eta}}}, p)
\]
